---
title: "Data Analysis Project"
author: "Katharine Wilson, Samuel Jane and Loretta Puwler"
date: "2024-11-16"
output: html_document
---
```{r}
install.packages("janitor")
```

```{r}
library(tidyverse)
library(tidycensus)
library(janitor)
library(dplyr)
```


Questions: 
1. What are the average ages of people in counties in Maryland with the highest rates of election day voting and mail-in voting? 
2. Which precincts have the highest rates of early voting of people who voted? 
3. How does early voting participation in Maryland for Governor compare to presidential elections? 
4. For the precincts that swayed the most towards Trump from 2020 to 2024, what is their average income? 
5. Did counties and precincts with heavy Black populations in Maryland shift toward Trump as much as other counties and areas did nationwide? Could compare PG and Baltimore to cities like Milwaukee, Detroit and others.


First we are setting up the data sets for use. 

Early voting data from 2024:
```{r}
early_voting <- read_csv("EarlyVoting RAW data.csv")
```

```{r}
md_earlyvote2024 <- early_voting |>
  mutate(across(starts_with("Day"), ~ replace_na(. , 0)))|>
  mutate(Early_Total2024 = Day1 + Day2 + Day3 + Day4 + Day5 +Day6 + Day7 + Day8)
```

```{r}
md_earlyvote2024_squeezed <- md_earlyvote2024 |>
  group_by(COUNTY_NAME)|>
  summarize(
  sum(Early_Total2024))
```

**IMPORTANT ONE 
```{r}
precincts_md_earlyvote2024 <- md_earlyvote2024 |> 
  group_by(COUNTY_NAME, Precinct)|>
  summarize(
    sum(Early_Total2024)
  )
```


```{r}
precincts_md_earlyvote2024 <- precincts_md_earlyvote2024 |>
  rename(Early_Total2024 =`sum(Early_Total2024)`)
```


Total voting data from 2024:
```{r}
total_election_2024 <- read_csv("PG24_AllPrecincts.csv")
```

```{r}
total_election_2024 <- total_election_2024 |>
mutate(Total2024 = `Early Votes` + `Election Night Votes` + `Mail-In Ballot 1 Votes` + `Mail-In Ballot 2 Votes` + `Provisional Votes`)
```

```{r}
total_election_2024_squeezed <- total_election_2024 |>
  group_by(`County Name`)|>
  summarize(
  sum(Total2024))
```

```{r}
precinct_total_election_2024 <- total_election_2024 |>
  group_by(`County Name`, `Election District - Precinct`)|>
  summarize(
  sum(Total2024))
```

```{r}
precinct_total_election_2024 <- precinct_total_election_2024 |>
  mutate(`County Name` = str_remove(`County Name`," County"))
```

```{r}
precinct_total_election_2024 <- precinct_total_election_2024 |>
  rename(Total2024 =`sum(Total2024)`)
```

//////////////////////////////////////////////////
Question #1:
What are the average ages of people in counties in Maryland with the highest rates of election day voting and mail-in voting? 

First, we need to find the counties with the highest rates of election day presidential voting and mail in voting in 2024. 

```{r}
voter_file_md <- read_csv("md_voters_for_match.csv")
```

```{r}
voting_counts <- total_election_2024 |> 
  mutate(election_day_votes = `Election Night Votes`)|>
  mutate(Mail_In = `Mail-In Ballot 1 Votes`)|>
  filter(`Office Name` == "President - Vice Pres")
```

```{r}
voting_count_sum <- voting_counts |>
  group_by(`County Name`)|>
  summarise(sum(Mail_In), sum(election_day_votes), sum(Total2024))|>
  mutate(MailInRate = (`sum(Mail_In)`/`sum(Total2024)`)*100)|>
  mutate(DayRate =(`sum(election_day_votes)`/`sum(Total2024)`)*100)
```

This shows the precincts with the highest rate of people who voted by mail (Howard, Anne Arundel, Calvert)
```{r}
top_mail24 <- voting_count_sum |>
  arrange(desc(MailInRate))
```

This shows the top precincts for people who voted on election day (Allegany, Caroline, Dorchester)
```{r}
top_day <- voting_count_sum |>
  arrange(desc(DayRate))
```

Now, we're opening the ACS data to find the average age in these counties. 
```{r}
acs_data_age <- get_acs( geography = "county", variables = "B07402_001", year = 2021, survey = "acs5", output = "wide")
```

For counties with the highest rates of mail-in voting, here are the average ages: 
- Howard 39.3
- Anne Arundel 39.3
- Calvert 40.6

For counties with the highest rates of election day voting, here are the average ages: 
- Allegany 42.3
- Caroline 39.9
- Dorchester 43.8 

These results show that counties with a higher rate of election-day voting had an older average population. This could hint that older people are more inclined to vote in the traditional election-day format. 

These results also how that counties with a higher rate of mail-in voters were younger than those with high election-day voting, implying that younger people in Maryland may be more open to voting by mail (a format promoted in more recent elections), 
////////////////////////////////////////////////

Question #2: 
Which precincts have the highest rates of early voting of people who voted? 

First, we are going to join the csvs for early voting and total voting in each precinct together. We are also replacing NAs with 0s. 
```{r}
combined_2024 <- precinct_total_election_2024 |>
  left_join(precincts_md_earlyvote2024, join_by(`County Name` == COUNTY_NAME, `Election District - Precinct` == Precinct))|>
  mutate(across(starts_with("Early"),~ replace_na(. , 0)))|>
  mutate(across(starts_with("Total"),~ replace_na(. , 0)))
```

Then, we are going to find the percent of people who voted early out of the total number of people who voted per precinct by creating a new column. 

```{r}
combined_2024 <- combined_2024 |>
  mutate(Early_Turnoutrate_2024 = (Early_Total2024/Total2024)*100)
```

Now, we are going to order them from highest early voting turnout to lowest. 
```{r}
combined_2024 <- combined_2024|>
  arrange(desc(Early_Turnoutrate_2024))
```

Our answer:
The precinct with the highest early voting turnout rate in Maryland was in Anne Arundel County (precinct 005-032). The rest of the top five precincts were from Wicomico and Somerset. 

According to census reproter, this precinct has an average age of 33 (with most people between 18 and 64). The average income per household is $108,000 and most people live in homes with 2.4 people inside. This implies that most people who live in this precinct are comfortable and of an age where getting transportation and heading to the polls early is easier. 

///////////////////////////////////////////////////////////////////////////////////////////////////////////////////////////
--------------
Question #3: 
How does early voting participation in Maryland for Governor compare to presidential elections? 

First, we are going to load in the data for early voting 2022. 

```{r}
early_voting_2022 <- read_csv("EarlyVoting RAW data 2022.csv")
```

```{r}
md_earlyvote2022 <- early_voting_2022 |>
  mutate(across(starts_with("Day"), ~ replace_na(. , 0)))|>
  mutate(Early_Total2022 = Day1 + Day2 + Day3 + Day4 + Day5 +Day6 + Day7 + Day8)
```

```{r}
md_earlyvote2022_squeezed <- md_earlyvote2022 |> 
  group_by(COUNTY_NAME)|>
  summarize(
  sum(Early_Total2022))
```

```{r}
md_earlyvote2022_squeezed <-md_earlyvote2022_squeezed |>
  rename(Early_Total2022 = `sum(Early_Total2022)`)
```

```{r}
md_earlyvote2024_squeezed<-md_earlyvote2024_squeezed|>
     rename(Early_Total2024 = `sum(Early_Total2024)`)
```

Now, we are loading in the total election vote records for 2022 and cleaning up the data. (this is the next few code blocks)
```{r}
total_vote_2022 <- read.csv("GG22_AllPrecincts.csv")
```

```{r}
total_vote_2022 <- total_vote_2022|>
mutate(Total2022 = Early.Votes + Election.Night.Votes + Mail.In.Ballot.1.Votes + Mail.In.Ballot.2.Votes + Provisional.Votes)
```

```{r}
total_vote2022_squeezed <- total_vote_2022 |>
  group_by(County.Name)|>
  summarize(
  sum(Total2022))
```

```{r}
total_vote2022_squeezed <- total_vote2022_squeezed|>
  rename(Total2022 = `sum(Total2022)`)
```

```{r}
total_vote2022_squeezed <-total_vote2022_squeezed|>
 mutate(County.Name = str_remove(County.Name," County"))
```

```{r}
md_earlyvote2022_squeezed <- md_earlyvote2022_squeezed|>
  mutate(COUNTY_NAME = str_remove(COUNTY_NAME," County"))
```


Now, we will make a csv that includes the total and early vote numbers for the 2022 election
```{r}
together_2022 <- total_vote2022_squeezed |>
  left_join(md_earlyvote2022_squeezed, join_by(County.Name == COUNTY_NAME))
```

Now, we will make a csv that includes the total and early vote numbers for the 2024 election

```{r}
total_election_2024_squeezed <- total_election_2024_squeezed |>
  rename(Total2024 = `sum(Total2024)`)
```

```{r}
total_election_2024_squeezed <- total_election_2024_squeezed |>
  mutate(`County Name` = str_remove(`County Name`," County"))
```

```{r}
md_earlyvote2024_squeezed <- md_earlyvote2024_squeezed |>
  mutate(COUNTY_NAME = str_remove(COUNTY_NAME," County"))
```

```{r}
together_2024 <- total_election_2024_squeezed |>
  left_join(md_earlyvote2024_squeezed, join_by(`County Name` == COUNTY_NAME))
```

```{r}
all_together <- together_2024 |> left_join(together_2022, join_by(`County Name` == County.Name))
```

```{r}
all_together <- all_together |>
  mutate(EarlyRate2022 = (Early_Total2022/Total2022)*100)|>
  mutate(EarlyRate2024 = (Early_Total2024/Total2024)*100)|>
  mutate(Rate_Diff_2422 = EarlyRate2024 - EarlyRate2022)|>
  arrange(desc(Rate_Diff_2422))
```

This shows that Charles County had the highest growth in early voting between 2022 and 2024. The county increased it's rate of voters choosing early voting by five percentage points. 

Charles County, according to census reporter, has a lower proportion of people below the poverty line compared to all of Maryland which could contribute to the population's ability to physically get to early voting and the opportnity to become more informed in state elections (because they likely don't have to worry about getting through the day as much). The population of Charles County is 50% Black, which could impact their incentives to vote early in national elections.
////////////////////////////////
Question #4: 
For the counties that swayed the most towards Trump from 2020 to 2024, what is their average income?  

First, we are going to load in the 2020 total voting data. 

```{r}
Total_2020 <-read_csv("All_By_Precinct_2020_General.csv")
```


  rename(TotalVotes ==`Total Votes`)

```{r}
Total_2020 <- Total_2020 |>
  filter(`Office Name` == "President - Vice Pres")|>
  rename(TotalVotes = `Total Votes`)|>
  mutate(across(starts_with("Total"), ~ replace_na(. , 0)))
```


```{r}
Trump_Total_2020 <- Total_2020 |>
  filter(`Candidate Name` == "Donald J. Trump")|>
  mutate(across(starts_with("Total"), ~ replace_na(. , 0)))
```

```{r}
Trump_Total_2024 <-total_election_2024|>
  filter(`Candidate Name` == "Donald J. Trump and JD Vance")|>
    mutate(across(starts_with("Total"), ~ replace_na(. , 0)))
```

This shows us how many people voted for Trump in each MD county in 2024: 
```{r}
Total24CountyTrump <- Trump_Total_2024 |>
  group_by(`County`)|>
  summarize(
  sum(Total2024))|>
  rename(Trump24 = `sum(Total2024)`)
```

This shows us how many people voted for Trump in each MD county in 2020: 
```{r}
Total20CountyTrump <- Trump_Total_2020 |>
  group_by(County) |>
  summarize(sum(TotalVotes))
```

This shows us how many people voted for any presidential candidate in 2020: 
```{r}
Total_2020_pres <- Total_2020 |>
  mutate(across(starts_with("Total"), ~ replace_na(. , 0)))|>
  group_by(County) |>
  summarize(sum(TotalVotes))
```

The second codeblock shows us the percent of presidential voters who voted for trump in each county in 2020, found after joining the two: 
```{r}
Trump20TotalCompare <- Total20CountyTrump |>
  left_join(Total_2020_pres, join_by(County == County))
```

```{r}
Trump20TotalCompare <- Trump20TotalCompare |>
  rename(Allpresvotes = `sum(TotalVotes).y`)|>
  rename(Trumpvotes = `sum(TotalVotes).x`)|>
  mutate(PercentTrumpWon = (Trumpvotes/Allpresvotes)*100)
```

```{r}
Total_2024_pres<- total_election_2024 |>
  filter(`Office Name` == "President - Vice Pres")|>
  mutate(across(starts_with("Total"), ~ replace_na(. , 0)))|>
  group_by(County)|>
  summarize(sum(Total2024))
```

```{r}
Total_2024_pres <- Total_2024_pres |>
  rename(Totalpres2024 = `sum(Total2024)`)
```

```{r}
Trump24TotalCompare <- Total_2024_pres |>
  left_join(Total24CountyTrump, join_by(County == County))
```

```{r}
Trump24TotalCompare <- Trump24TotalCompare |>
  mutate(PercentTrumpWon24 = (Trump24/Totalpres2024)*100)
```

This will bring all the numbers together:
```{r}
AllTrumppres2420 <-Trump24TotalCompare |>
  left_join(Trump20TotalCompare, join_by(County == County))
```

```{r}
AllTrumppres2420 <-AllTrumppres2420 |>
  mutate(Diff2420 = PercentTrumpWon24 - PercentTrumpWon)|>
  arrange(desc(Diff2420))
```

The code above shows that the top counties that leaned more towards Trump in 2024 than in 2020 were Montgomery, Calvert and Caroline counties. 

Now, we're going to look at what each of these counties average income is. 
//////////
```{r}
census_api_key("41f507ff77bfed12e56b9c9420926d8c02d5b09d", install=TRUE)
```

```{r}
acs_data <- get_acs( geography = "county", variables = "B19013_001", year = 2021, survey = "acs5", output = "wide")
```

^ Looking at this census data, which shows average income in 2021 in 2023 data the counties have the following average incomes: 
Montgomery County: $117,345
Calvert County: $120,295 per year
Caroline County: $63,027 per year

These average incomes show that there was a broad switch to favoring Trump that included a wide range of average county wages. Montgomery and Calvert counties both have a relatively high average income were the top counties who swayed to Trump. Caroline County, in third place, had half of the average income as the top-two counties. 

The counties that swayed the least towards Trump were Garrett, Charles and Anne Arundel. These counties had average incomes of $58,011 (Garrett), $107,808 (Charles) and $108,048. This shows again that the statwide lean towards Trump in Maryland counties did not seem to be determined by average income level - even though the economy was a big issue in campaign messaging. 

```


acs_data <- get_acs( geography = "county",  # You can change this to "state" or "tract" if needed variables ="B19013_001",  # Median household income variable code year = 2021,  # Year of ACS data survey = "acs5",  # 5-year estimates output = "wide"  # Wide format (gives you 'estimate' and 'moe' columns) ) # Check the structure of the datastr(acs_data) # Clean and calculate the average income by county average_income_by_county <- acs_data %>%select(NAME, estimate) %>%  # Keep only county name and income estimate group_by(NAME) %>%  # Group by county name summarize(avg_income = mean(estimate, na.rm = TRUE))  # Calculate average income per county # View the results head(average_income_by_county)

Now that we have our data sets that include the number of people who voted for president in each county and the number

NOTE: 
01: Allegany
02: Anne Arundel 
03: Baltimore City 
04: Baltimore
05: Calvert
06: Caroline
07: Carroll
08: Cecil
09:Charles
10: Dorchester
11:Frederick 
12: Garrett
13: Harford
14: Howard
15: Kent
16: Montgomery 
17: Prince George's 
18: Queen Anne's 
19: St. Mary's
20: Somerset
21: Talbot
22: Washington 
23: Wicomico 
24: Worchester 
Then we will load the census data for the largest shifts and see of the average income correlates.

//////////////////////////////////////////////////////////////////////////
Question #5: Did counties and precincts with heavy Black populations in Maryland shift toward Trump as much as other counties and areas did nationwide? Could compare PG and Baltimore to cities like Milwaukee, Detroit and others.

We will be using census racial demographics for this (the acs5)

```{r}
acs5 <- load_variables(2022, "acs5", cache = TRUE)
View(acs5)
```

B02001_003 (Black or African American alone)

We will load the census data for maryland with the highest black populations and then join that with the Trump voter percentage by county. Then we will add the raw number for Trump percentage and then compare it to maryland

------NOTES______________

Loading in census data 


We will have to merge the 5-year acs to: 
Find the average income for question 4 
Find the precincts with the highest Black population for question 5 

We will have to also use the 2020 early voting data and 2022 data for questions 3 and 4. 

