---
title: "lab_02"
author: "Katharine Wilson"
date: "2024-09-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

For this lab we'll be working with a slice of Maryland voter registration data: records of people who have registration dates this year through Aug. 7. Let's load it.

```{r}
md_voters_2024 <- read_rds("data/lab_02.rds")
```

## Examine Data

Take a look at the data, getting familiar with what it contains. Most of the columns should be self-explanatory, but there are two date columns: STATE_REGISTRATION_DATE and COUNTY_REGISTRATION_DATE. The former represents the date the person registered as a Maryland voter; the latter is when they registered in their current county. They may be the same, but if a voter moves from one county to another they will differ. This is important because if the two dates are the same, that's a pretty good indication that this is a *new* voter, someone who has no history of registering in Maryland. You'll need to keep that definition in mind.

## Answer questions

**Q1** How many voters appear to be *new* registrants since July 21, when President Joe Biden announced he would leave the race, based on the information above? Create a dataframe to hold the only those voters. Which counties have seen the most? Take a look at county population figures here and compare the two in your answer: <https://worldpopulationreview.com/us-counties/maryland>.

Montgomery County and Baltimore City had the most new registrations since Biden dropped out of the race. Montgomery County has the largest population of any county in Maryland and the city is the 5th most-populous county. These large populations contributed to the higher number of new registrations compared to other counties. 

```{r}
new_md_voters_2024 <- md_voters_2024 |> filter (STATE_REGISTRATION_DATE > "2024-07-21") |> group_by(COUNTY) |> summarize(
  voters_in_county = n()
) |>
arrange(desc(voters_in_county))
```

**Q2** A better question is whether the new voters you found in Q1 are an outlier compared to other weeks this year. Using the original dataframe with 2024 registrants, use `mutate()` and lubridate to produce week numbers from the COUNTY_REGISTRATION_DATE (Lubridate has a function that turns a date into a week number. Use that to answer this question). Save that to the original dataframe. Then calculate how many voters are in each week. Describe the distribution of voters groups in a sentence - are the weeks from July 21-August 7 different from the rest, and if so, how?

The weeks after Biden dropped out (July 21 - August 7, aka weeks 30 and 31 ) are among the biggest weeks for Maryland registrations - with the week of July 22 having the second largest number of registrations. However, the week with the most voter registrations was in early June before the first presidential debate (week 24). Also among weeks with the highest numbers of registrations were the weeks after the first debate and before Biden dropped out. Overall, the trends seem to indicate that a lot of people registered to vote this summer. A lot of major events happened this summer - including the downfall of the Biden campaign, the debates and the RNC - which each likely also impacted registrations. Therefore, there was a boost after Biden dropped out but this isn't very different from the boost in registrations throughout the summer. 

```{r}
md_voters_2024 |>
  mutate(week_number = isoweek(COUNTY_REGISTRATION_DATE)) |> 
  group_by(week_number) |>
  summarize(
    voters_per_week = n()) |>
  arrange(desc(voters_per_week))
```

**Q3** Using lubridate, show which month saw the largest number of registrations. Does the answer help shed more light on Q2?

This helps determine what months have the most on-average voter registrations. While July does have the most voter registrations, possibly due to Biden dropping out, it was a part of an existing pattern of a lot of people registering to vote this summer. The second-highest month for voter registrations was June, a month unaffected by Biden ending his reelection bid. 
```{r}
md_voters_2024 |>
  mutate(month_number = month(COUNTY_REGISTRATION_DATE)) |> 
  group_by(month_number) |>
  summarize(
    voters_per_month= n()) |>
  arrange(desc(voters_per_month))
```

**Q4** Maryland has, officially, five regions: <https://www.visitmaryland.org/article/maryland-regions>. Modify the original dataframe again to add a `region` column based on the `COUNTY` value using case_when. Are you going to have to do this for every Maryland county plus Baltimore City? Yes, you are. Which regions and parties account for the most new voters?

The most new voters came from Central Maryland and the most amount of new voters registered with the Democratic Party. 
```{r}
### What regions account for the most new voters
md_voters_regions <- md_voters_2024 |> filter (STATE_REGISTRATION_DATE == COUNTY_REGISTRATION_DATE)|> mutate(region = case_when(
                         COUNTY == "Garrett" ~ "Western Maryland",
                         COUNTY == "Allegany" ~ "Western Maryland",
                         COUNTY == "Washington" ~ "Western Maryland",
                         COUNTY == "Frederick" ~ "Capital Region",
                         COUNTY == "Montgomery" ~ "Capital Region",
                         COUNTY == "Prince George's" ~ "Capital Region",
                         COUNTY == "Anne Arundel" ~ "Central Maryland",
                         COUNTY == "Baltimore City" ~ "Central Maryland",
                         COUNTY == "Baltimore County" ~ "Central Maryland",
                         COUNTY == "Carroll" ~ "Central Maryland",
                         COUNTY == "Harford" ~ "Central Maryland",
                         COUNTY == "Howard" ~ "Central Maryland",
                         COUNTY == "Calvert" ~ "Southern Maryland",
                         COUNTY == "Charles" ~ "Southern Maryland",
                         COUNTY == "Saint Mary's" ~ "Southern Maryland",
                         COUNTY == "Kent" ~ "Eastern Shore Maryland",
                         COUNTY == "Queen Anne's" ~ "Eastern Shore Maryland",
                         COUNTY == "Talbot" ~ "Eastern Shore Maryland",
                         COUNTY == "Caroline" ~ "Eastern Shore Maryland",
                         COUNTY == "Dorchester" ~ "Eastern Shore Maryland",
                         COUNTY == "Wicomico" ~ "Eastern Shore Maryland",
                         COUNTY == "Somerset" ~ "Eastern Shore Maryland",
                         COUNTY == "Worcester" ~ "Eastern Shore Maryland",
                         COUNTY == "Cecil" ~ "Eastern Shore Maryland",
                         )) |>
  group_by(region)|>
  summarise(
    voters= n())|>
  arrange(desc(voters))

### What parties account for the most new voters
md_voters_parties  <- md_voters_2024 |> filter (STATE_REGISTRATION_DATE == COUNTY_REGISTRATION_DATE)|> group_by(PARTY)|>
  summarise(
    voters= n())|>
  arrange(desc(voters))
```

**Q5** What's an additional question you could ask of this data?

I could ask what party on-average was getting the most new registrations before Biden dropped out vs. after Biden dropped out. 
