---
title: "Class Reference"
author: "Katharine Wilson"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Consider this a personal guide to the commands and functions you will learn. In general, when you come across an R command or function that you want to remember, put it in here along with a description of what it does and when you'd use it.

### How to set the working directory

The command to set a working directory is setwd(). For example, this sets the working directory to a directory called data_journalism_2024_fall inside a directory called jour472 located inside your home directory (the ~ is equivalent of /Users/[youruser] on a Mac).

```{r}
setwd("~/jour472/data_journalism_2024_fall")
```

### How to install and load the tidyverse

```{r}
install.packages("tidyverse")
library(tidyverse)
```

### How to name variables
```{r}
example <-2
```

### How to import data 
```{r}
variable_name <- read_rds("datatable_from_rdsfile")
```

### How to see a list of columns, data type per column and first few values
```{r}
glimpse(variable_name)
```

### How to count # of contributions
```{r}
variable_name |>
  group_by(memo_text)|>
  summarize(count_contribs = n()
            )
```

### How to order something (decending)
```{r}
  arrange(desc(total_amount))
```
### Order accending
```{r}
  arrange(total_amount)
```

### How to show min/max/median/mean
```{r}
wv_summary <-westvirginia_winred_contributions |>
  group_by(memo_text) |>
  summarise(
    count_contribs = n(),
    total_amount = sum(amount),
    mean_amount = mean(amount),
    median_amount = median(amount),
    min_amount = min(amount),
    max_amount = max(amount)
  ) |>
  arrange(desc(total_amount))
```
```
### Changing the origional data
```{r}
texas_precinct_20 <- texas_precinct_20 |>
  mutate(election_day = case_when(
    election_day == '5+' ~ '49',
    .default = election_day
  ))
  
  ```
  
```{r}

```
  

### Cleaning names of columns 
```{r}
cleaned_conowingo <- conowingo |>
  clean_names()

```
### Rename a column 
```{r}
cleaned_conowingo <- conowingo |>
  clean_names() |> 
  rename(linenumber = x1_linenumber)
```
### Removing duplicates
```{r}
cleaned_conowingo <- conowingo |>
  clean_names() |> 
  rename(linenumber = x1_linenumber) |> 
  mutate(amount = as.numeric(amount)) |> 
  distinct()
```
### Correct mispellings 
```{r}
cleaned_conowingo <- conowingo |>
  clean_names() |> 
  rename(linenumber = x1_linenumber) |> 
  mutate(amount = as.numeric(amount)) |> 
  distinct() |>
  mutate(zip = str_sub(zip, start=1L, end=5L)) |>
  mutate(city = str_to_title(city)) |>
  mutate(city = case_when(
    str_detect(city,"^Conowing") ~ "Conowingo",
    TRUE ~ city
  ))
```

