---
title: "lab_05"
author: "Katharine Wilson"
date: "2024-10-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# load the tidyverse
options(scipen=999)
library(tidyverse)
library(lubridate)
```

## Getting Started

This week we'll be combining and joining data to make it more useful and to ask some questions about it. This week, we'll revisit our new Maryland voters data and try to join it with contributions to the Harris campaign from Maryland donors made via ActBlue. Can we find new voters who are not registered Democrats who gave to Harris after she became a presidential candidate on July 21? What, if anything, can we say about those people?

We'll be doing something similar to what Ence Morse did for [this Washington Post story](https://www.washingtonpost.com/politics/2024/09/24/kamala-harris-younger-donors/?pwapi_token=eyJ0eXAiOiJKV1QiLCJhbGciOiJIUzI1NiJ9.eyJyZWFzb24iOiJnaWZ0IiwibmJmIjoxNzI3MzIzMjAwLCJpc3MiOiJzdWJzY3JpcHRpb25zIiwiZXhwIjoxNzI4NzA1NTk5LCJpYXQiOjE3MjczMjMyMDAsImp0aSI6ImIwYWMxNTk4LWMwMDgtNDE5Ny04MjAyLWVmMWNjZWFmYWZiMiIsInVybCI6Imh0dHBzOi8vd3d3Lndhc2hpbmd0b25wb3N0LmNvbS9wb2xpdGljcy8yMDI0LzA5LzI0L2thbWFsYS1oYXJyaXMteW91bmdlci1kb25vcnMvIn0.3-s0tBkMV7Y9DD21CBsKMlaliGlOKT2egCO-UFT735Y), just on a smaller scale.

First, load two CSV contribution files in the `data` folder, one each for donations made in July and August, and then combine them into a single dataframe.

```{r}
# load the Harris contribution data files and combine them
july_harris <- read_csv("data/july_harris.csv")
august_harris <- read_csv("data/august_harris.csv")

harris_contribs_combined <- bind_rows(list(july_harris, august_harris))
View(harris_contribs_combined)
```

Next, let's read in the new voters (people who registered in 2024):

```{r}
# load the new voters data
new_voters <- read_csv("data/new_voters.csv")
```

Next join the Harris donations to the voter data. Remember, people's names are common, __so you'll want to use name columns and address information__. You'll need to figure out which columns to use from both dataframes. The new_voters data has both a single Address column and various parts of the address, such as HouseNumber. Make a plan for picking your join columns - try out some combinations - and save the results to a new dataframe.

```{r}
# join the combined Harris file to the new voters
# you will need to determine what columns to use
harris_contribs_and_new_voters <- new_voters |> inner_join(harris_contribs_combined, join_by(FirstName == first_name, LastName == last_name, ResidentialZip == zip, ResidentialCity == city, Address == address_one))
```

## Answer questions

Q1. Explain why you used the columns you did to make the join, and why that result is better than other ways you tried. Are there any problems with the join you made, such as the possibility of false positives?

A1. I used first name, last name, zip, city and address because that gave me the most specific match possible. These filters limit almost all false positive matches by limiting all matches to people in the same home with the same name. There could be a false positive if two people in the same house have the same name. We are also excluding people who's address is written differently in the two data sets (like rd and road or dr and drive). 

Q2. Based on your join, how many donations to Harris have come from people who registered this year? How many unique people does that represent, and which donors have given the largest _number_ of donations since July 20? 

A2. 
My join shows that 237 new voters donated to Harris in July and August. There were 324 individual donations, totaling $19,190. Sophia Hoffer-Perkins, Eric Snipe, Brittany Musolino were tied for the largest number of donations (7). 

```{r}
contribs_by_new_voter <- harris_contribs_and_new_voters |>
  group_by(VTR_ID, FirstName, LastName)|>
  summarise(
    count=n()
    )|>
arrange(desc(count))

harris_contribs_and_new_voters |> 
  summarise(
    sum(amount))
```

Q3. What are the party and gender breakdowns of those donors? How many donations has Harris gotten from non-Democrats?

A3. The party breakdown is as follows: 292 Democrats, 23 unaffiliated, 4 Republicans, 2 Libertarians, 2 others and 1 No Labels Marylanders. Harris therefore recived donations from 32 non-Democrats. The gender breakdown is as follows: 172 females, 140 males and 12 others. 

```{r}
harris_contribs_and_new_voters |>
  group_by(Party)|>
  summarise(
    count=n()
    )|>
arrange(desc(count))
```
```{r}
23+4+2+2+1 
```
```{r}
harris_contribs_and_new_voters |>
  group_by(Gender)|>
  summarise(
    count=n()
    )|>
arrange(desc(count))
```


Q4. Choose your own adventure: I want you to come up with a question you could ask of one of the dataframes we've created for this lab. It could involve just the new voters, or the donors or the joined dataframe, but it cannot be a variation of a previous question (so, for example, you can't ask, "How many Republicans have given to Harris"). First write the question and then try to answer it by writing code. If you're not sure of the exact code, try googling (or ask me for advice) but, at a minimum, describe what you think you'd have to do to answer it and what the results would look like. That can help.

A4: How many new voters who were younger than 30 years old on July 21, 2024 donated to Harris? 
45 new voters under 30 on July 21, 2024 donated to Harris. 
```{r}
harris_contribs_and_new_voters |>
  filter(BirthDate > "1994-07-21")
```

