---
title: "lab_04"
author: "Katharine Wilson"
date: "2024-09-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings

You'll need to load three packages for this: the tidyverse, lubridate and janitor.

**Task** load these three packages.

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(janitor)
library(lubridate)

```

For this lab, we want to investigate spending by the two leading Senate candidates in Maryland during this election, Angela Alsobrooks and Larry Hogan. Our goal is to try and standardize the purpose of the expenditures so that we can compare the two campaigns. We have the data, but it's messy - purpose descriptions are spelled differently and capitalized differently - and we need to clean it up. We'll use a combination of RStudio and OpenRefine to do that.

The basic process here is to start in RStudio, export data so that we can use OpenRefine to clean it up, and then bring it back into RStudio to finish our analysis.

## Load Data

You'll need to load one data set, a CSV file of campaign expenditures located in the data folder called "md_senate_expenses.csv"

**Task** Create a codeblock below, then read the data in and assign it to an appropriate variable name. You'll want to make sure that any date columns are actually date datatypes.

```{r}
# Load the Maryland expenditures data table
senate_expenses <- read_csv("data/md_senate_expenses.csv")
```

## Answer questions

**Q1.** You've been assigned to compare the spending priorities of the two campaigns, but they describe their spending differently using the `purpose` column. Using OpenRefine, create a project using the original CSV file, make a copy of the `purpose` column called `purpose_clean` and then standardize it, focusing on the purposes that appear most often or have the largest total amounts. You can transform the data as you want - making all records the same case, for example, to help. The goal is to be able to group together as many similar purpose values, so you can choose to call all transportation-related expenses the same thing.

How do you know when you're done? It's a judgment call, but you should be close to 100 unique values for `purpose_clean`, and maybe even below that number.

Then, when you are done standardizing `purpose_clean`, export it as a new CSV file to your data folder, *giving it a different name* and read it back into RStudio as a new dataframe.

Then, using that `purpose_clean` column, write code to find the largest amounts of spending on categories by both campaigns.

**A1.**
The largest categories the Alsobrooks campaign spent money on were advertising, staffing and list acquisition. The Hogan campaign spent the most money on advertising, staffing and fundraising services. 
```{r}
senate_expenses_clean <- read_csv("data/md_senate_expenses_OR_csv.csv")

senate_expenses_clean|>
  filter(candidate=="Alsobrooks")|>
  group_by(purpose_clean)|>
  summarise(
        amount_per_category = sum(amount)
  )|>
  arrange(desc(amount_per_category))
```
```{r}
senate_expenses_clean|>
  filter(candidate=="Hogan")|>
  group_by(purpose_clean)|>
  summarise(
        amount_per_category = sum(amount)
  )|>
  arrange(desc(amount_per_category))
```


**Q2.** Let's do some research on the top purposes from A1. Feel free to search the Web to understand those terms better if need be, and then write a description comparing and contrasting the two campaigns' largest categories of spending.

**A2.**
Both campaigns spent the most money on advertising (1) and staffing (2). Alsobrooks spent more than Hogan in both categories. The Alsobrooks campaign spent a lot of money on list acquisition - an attempt to find the best possible voters and donors through email. This reflects on Alsobrooks' need to get her name out there in a tight race against a well-known former governor. Hogan spent a significant amount of money on fundraising services. 

**Q3.** Pick one of the top categories for each campaign and write code that shows how that spending has changed over time, if at all. You'll need to use lubridate for this. Write a couple of sentences describing that pattern.

**A3.**
First, I looked at the list aquisition cost for the Alsobrooks campaign. There were only seven payments towards this category, mostly focused on late May and early June. This was after the primary, when the Alsobrooks campaign had to start turning their attention towards Larry Hogan and away from priamry competitor Rep. David Trone. At this point in the race, the campaign was likely trying to attract Democratic donors who are more likely to donate now that there is a Democratic nominee. Alsobrooks needed name recognition and the ability to reach out to new donors, thus, they needed a list of good people to contact.  

For the Hogan campaign, I looked into how his advertising spending changed over time. Hogan focused his advertising spending right before the Senate primary and just after the primary once it was clear he was the Republican nominee. Hogan, as a well-known former governor, doesn't have to spend as much time getting his name recognition out there, which could explain his occasional gaps in advertising. Instead, his campaign is spending their money letting Marylanders know that their former governor is running instead of the having to explain who he is. 
```{r}
senate_expenses_clean|>
  filter(candidate=="Alsobrooks", purpose_clean=="List Acquisition")|>
  mutate(week_number =  isoweek(date))|>
  group_by(week_number)|>
  summarise(
     amount_per_month = sum(amount)
  )|>
  arrange(desc(week_number))
```
```{r}
senate_expenses_clean|>
  filter(candidate=="Hogan", purpose_clean=="Advertising")|>
  mutate(week_number =  isoweek(date))|>
  group_by(week_number)|>
  summarise(
     amount_per_month = sum(amount)
  )|>
  arrange(desc(amount_per_month))
```

**Q4.** Which committee has the larger payroll or salary spending so far? How many unique recipients of payroll/staff spending are there for each candidate?

**A4.**
The Alsobrooks campaign has spent more on payroll/salary than the Hogan campain. There are 325 unique recepts for payroll/staff spending for Alsobrooks and 23 for Hogan.
```{r}
senate_expenses_clean|>
  filter(candidate=="Alsobrooks", purpose_clean=="Staffing")|>
  group_by(purpose_clean)|>
  summarize(
         amount_purpose = sum(amount)
  )
```
```{r}
senate_expenses_clean|>
  filter(candidate=="Alsobrooks", purpose_clean=="Staffing")
```
```{r}
senate_expenses_clean|>
  filter(candidate=="Hogan", purpose_clean=="Staffing")|>
  group_by(purpose_clean)|>
  summarize(
         amount_purpose = sum(amount)
  )
```
```{r}
senate_expenses_clean|>
  filter(candidate=="Hogan", purpose_clean=="Staffing")
```



**Q5** What's the most interesting story idea you've found in this data?

**A5**
I think it would be interesting to do a story on how much money the two campaigns spent on events (Alsobrooks spend more than double what Hogan spent) on event planning and what these events look like. Or, a broader story about how much Alsobrooks has had to spend to get name recognition. 

I would also want to do a story on where big purchases - like the Alsobrooks campaign's purchase of a car - go after the election. 
